High-Dimensional Regularized Discriminant Analysis

### 2016-01-13 reports/paper-hdrda.Rnw
Added DMY's changes.

\citeauthor{Friedman:1989tm} proposed the popular regularized discriminant
analysis (RDA) classifier that utilizes a biased covariance-matrix estimator to
partially pool the sample covariance matrices from linear and quadratic
disriminant analysis before shrinking the resulting estimator towards a scaled
identity matrix. The RDA classifier's two tuning parameters are typically
estimated via a computationally burdensome cross-validation procedure from a
grid of candidate values. We reformulate the RDA classifier in the small-sample,
high-dimensional setting and then show that the classification decision rule is
equivalent to a classifier in a subspace having a much lower dimension. As a
result, the dimension reduction involved yields a substantial reduction in
computation during model selection. Our parameterization also offers
interpretability previously lacking with the RDA classifier. We demonstrate with
three artificial and four real high-dimensional data sets that our proposed
classifier is often superior to several recently proposed sparse and regularized
classifiers in terms of classification accuracy. Finally, we provide an
implementation of our proposed classifier in the {\tt sparsediscrim} R package,
available on CRAN.

@@

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers}
employed in a variety of disciplines, such as medical research, facial
recognition, remote sensing, and chemical analysis
\citep*{Ching:2012fu,Li:2012ev,Zhang:2010va}. Despite the advances in
classification to facilitate technological advancements, well-established
classifiers often degrade in accuracy with modern, high-dimensional data sets
when the number of features $p$ exceeds the total sample size $N$, due to the
\emph{curse of dimensionality} \citep{Bellman:1961tn}. For instance, linear
discriminant analysis (\emph{LDA}) and quadratic discriminant analysis
(\emph{QDA}) are popular because of their simplicity, speed of computation, and
interpretability. However, both methods are incalculable when $p > N$ because
the class sample covariance matrices are singular \citep*{Bouveyron:2007gx}.

@@

Here, we present a new classification method intended for the $p > N$ case that
builds on the classic regularized discriminant analysis (\emph{RDA}) proposed by
\cite{Friedman:1989tm}. Our proposed high-dimensional \emph{RDA} (\emph{HDRDA})
classifier employs a biased covariance matrix estimator that partially pools the
individual sample covariance matrices from the \emph{QDA} classifier with the
pooled sample covariance matrix from the \emph{LDA} classifier. We then shrink
the resulting covariance-matrix estimator towards a scaled identity matrix to
ensure positive definiteness. By reparameterizing the \emph{RDA} classifier
similar to \cite*{Hastie:2008dt} and \cite{Halbe:2007ea}, we first show that the
pooling parameter in the \emph{HDRDA} classifier determines the contribution of
each training observation to the estimation of each class covariance matrix,
enabling interpretability that has been previously lacking with the \emph{RDA}
classifier \citep{Bensmail:1996fw}.

Next, we achieve a substantial reduction in dimension to reduce greatly the
computation during model selection by applying a reasoning similar to
\cite{Ye:2006tq}. We establish that the matrix operations corresponding to the
null space of the pooled sample covariance matrix are redundant and can be
discarded from the \emph{HDRDA} decision rule without loss of classificatory
information. Moreover, because the \emph{HDRDA} decision rule is invariant to
adjustments to the zero eigenvalues, the decision rule in the original feature
space is equivalent to a decision rule in a lower dimension, such that matrix
inverses and determinants of relatively small matrices can be rapidly computed.
Furthermore, we show that several shrinkage methods that are special cases of
the \emph{HDRDA} classifier have no effect on the zero eigenvalues of the
covariance-matrix estimators. Such methods include work from
\cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier when $p \gg N$. We then study the classification
performance of the \emph{HDRDA} classifier on four real high-dimensional data
sets along with a simulation design that generalizes the experiments initially
conducted by \cite{Guo:2007te}. We demonstrate that the \emph{HDRDA} classifier
often attains superior classification accuracy to several recent classifiers
designed for small-sample, high-dimensional data from \cite*{Tong:2012hw},
\cite{Witten:2011kc}, \cite*{Pang:2009ik}, and \cite{Guo:2007te}. We also
include as a benchmark the random forest from \cite{Breiman:2001fb} because
\cite*{FernandezDelgado:2014ul} has concluded that the random forest is often
superior to other classifiers in benchmark studies.

The remainder of this document is organized as follows. In Section
\ref{sec:preliminaries} we present the discrimination problem and the necessary
notation to describe our contributions. In Section \ref{sec:rda} we present the
\emph{HDRDA} classifier along with its interpretation, properties, and
computationally efficient model-selection procedure. Section \ref{sec:sims}
describes our simulation studies of artificial and real data sets and also
describes the experimental results. We conclude with a brief discussion in
Section \ref{sec:discussion}.

@@

The smallest eigenvalues of $\widehat{\bm \Sigma}_k$ and the directions
associated with their eigenvectors can highly influence \eqref{eq:qda}. In fact,
the eigenvalues of $\widehat{ \bm \Sigma}_k$ are well-known to be biased if $p
\ge n_k$ such that the smallest eigenvalues are underestimated
\citep{Seber:2004uh}. Moreover, if $p > n_k$, then rank$(\widehat{\bm \Sigma}_k)
\le n_k$, which implies that at least $p - n_k$ eigenvalues of $\widehat{\bm
  \Sigma}_k$ are zero. Furthermore, although more feature information is
available to discriminate among the $K$ classes, if $p > n_k$, \eqref{eq:qda} is
incalculable because $\widehat{\bm \Sigma}_k^{-1}$ does not exist.

Several regularization methods, such as the methods considered by
\cite{Xu:2009fl}, \cite{Guo:2007te}, and \cite{Mkhadri:1995jp} have been
proposed in the literature to adjust the eigenvalues of $\widehat{\bm \Sigma}_k$
so that \eqref{eq:qda} is calculable and has reduced variability in the standard
bias-variance tradeoff. A common form of the covariance-matrix regularization
applies a shrinkage factor $\gamma > 0$, so that
\begin{align}
	\widehat{\bm \Sigma}_k(\gamma) = \widehat{\bm \Sigma}_k + \gamma \bm I_p, \label{eq:ridge-estimator}
\end{align}
similar to a method employed in ridge regression \citep{Hoerl:1970cd}. Equation
\eqref{eq:ridge-estimator} effectively \emph{shrinks} the sample covariance
matrix $\widehat{\bm \Sigma}_k$ toward $\bm I_p$, thereby increasing the
eigenvalues of $\widehat{\bm \Sigma}_k$ by $\gamma$. Specifically, the zero
eigenvalues are replaced with $\gamma$, so that \eqref{eq:ridge-estimator} is
positive definite. For additional covariance-matrix regularization methods, see
\cite{Ramey:2013ji}, \cite{Xu:2009fl}, and \cite{Ye:2009gd}.

@@

Here, we define the \emph{HDRDA} classifier by first formulating the
covariance-matrix estimator $\widehat{\bm \Sigma}_k(\lambda)$ and demonstrating
its clear interpretation as a linear combination of the crossproducts of the
training observations centered by their respective class sample means. We define
the convex combination
\begin{align}
  \widehat{\bm \Sigma}_k(\lambda) \coloneqq (1 - \lambda) \widehat{\bm \Sigma}_k + \lambda
  \widehat{\bm \Sigma}, \quad k = 1, \ldots, K,\label{eq:sig-lambda-alternative}
\end{align}
where $\lambda \in [0, 1]$ is the \emph{pooling} parameter. By rewriting
\eqref{eq:sig-lambda-alternative} in terms of the observations $ \bm x_i$ $(i =
1, \ldots, N)$, each centered by its class sample mean, we attain a clear
interpretation of $\widehat{\bm \Sigma}_k(\lambda)$. That is,
\begin{align}
	\widehat{\bm \Sigma}_k(\lambda)
  &= \left( 1 - \lambda + \frac{\lambda n_k}{N} \right) \widehat{\bm \Sigma}_k +  \frac{\lambda}{N} \sum_{\substack{k' = 1\\k' \ne k}}^K n_{k'} \widehat{\bm \Sigma}_{k'} \nonumber \\
	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right)\sum_{i=1}^N
  I(y_i = k) \bm x_i \bm x_i^{T} +  \frac{\lambda}{N} \sum_{i=1}^N I(y_i \ne k)
  \bm x_i \bm x_i^{T} \nonumber \\
	&= \sum_{i=1}^N c_{ik}(\lambda) \bm x_i \bm x_i^{T},\label{eq:sig-lambda-alternative2}
\end{align}
where $c_{ik}(\lambda) = \lambda N^{-1} + (1 - \lambda)n_k^{-1}I(y_i = k)$.
From \eqref{eq:sig-lambda-alternative2}, $\lambda$ weights the contribution of
each of the $N$ observations in estimating $\bm \Sigma_k$ from all $K$ classes
rather than $n_k$ observations from a single class. As a result, we can
interpret \eqref{eq:sig-lambda-alternative2} as a covariance-matrix estimator
that borrows from $\widehat{\bm \Sigma}$ in \eqref{eq:pooled-cov} to estimate $
\bm \Sigma_k$.

Below, we will show that the pooling operation is advantageous in increasing the
rank of each $\widehat{\bm \Sigma}_k(\lambda)$ from rank$(\widehat{\bm
  \Sigma}_k)$ to rank$(\widehat{\bm \Sigma})$ for $0 < \lambda \le 1$. Notice
that if $\lambda = 0$, then the observations from the remaining $K - 1$ classes
do not contribute to the estimation of $\bm \Sigma_k$, corresponding to
$\widehat{\bm \Sigma}_k$. Furthermore, if $\lambda = 1$, the weights
$c_{ik}(\lambda)$ in \eqref{eq:sig-lambda-alternative2} reduce to $1/N$,
corresponding to $\widehat{\bm \Sigma}$. For brevity, when $\lambda = 1$, we
define $\bm X = [\sqrt{c_{1k}(1)} \bm x_1^{T}, \ldots, \sqrt{c_{Nk}(1)} \bm
  x_N^{T}]^{T}$ such that $\widehat{\bm \Sigma} = N^{-1} \bm X^{T} \bm X$. Similarly,
for $\lambda = 0$, we define $\bm X_k = [\sqrt{c_{1k}(0)} \bm x_1^{T}, \ldots,
  \sqrt{c_{Nk}(0)} \bm x_N^{T}]^{T}$ such that $\widehat{\bm \Sigma}_k = n_k^{-1} \bm
X_k^{T} \bm X_k$. In Figure \ref{fig:hdrda-contours} we plot the contours of
five multivariate normal populations for $\lambda = 0$ with unequal covariance
matrices. As $\lambda$ approaches 1, the contours become more similar, resulting
in identical contours for $\lambda = 1$.

@@

As we have discussed above, several eigenvalue adjustment methods have been
proposed to increase eigenvalues (approximately) equal to 0. To further improve
the estimation of $\bm \Sigma_k$ and to stabilize the estimator's inverse, we
define the eigenvalue adjustment of \eqref{eq:sig-lambda-alternative} as
\begin{align}
	\tilde{\bm \Sigma}_k \coloneqq \alpha_k \widehat{\bm \Sigma}_k(\lambda) + \gamma \bm I_p,\label{eq:hdrda-cov}
\end{align}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. Thus, the \emph{pooling} parameter $\lambda$ controls the amount
borrowed from $\widehat{\bm \Sigma}$ to estimate $\bm \Sigma_k$, and the
\emph{shrinkage} parameter $\gamma$ determines the degree of eigenvalue
shrinkage. The choice of $\alpha_k$ allows for a flexible formulation of
covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1, \ldots,
K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}. Similarly,
if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form similar to the
\emph{RDA} classifier from \cite{Friedman:1989tm}. Substituting
\eqref{eq:hdrda-cov} into \eqref{eq:qda}, we define the \emph{HDRDA} classifier
\begin{align}
	D_{HDRDA}(\bm x) = \argmin_{k}  (\bm x - \xbar_k)^{T}\tilde{\bm \Sigma}_k^{+}(\bm x - \xbar_k)  + \log |\tilde{\bm \Sigma}_k|. \label{eq:hdrda}
\end{align}
For $\gamma > 0$, $\tilde{\bm \Sigma}_k$ is nonsingular such that $\tilde{\bm
  \Sigma}_k^{-1}$ can be substituted for $\tilde{\bm \Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\bm \Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\bm
  \Sigma}_k$. Following \cite{Friedman:1989tm}, we select $\lambda$ and $\gamma$
from a grid of candidate models via cross-validation. We provide an
implementation of \eqref{eq:hdrda} in the {\tt hdrda} function in the {\tt
  sparsediscrim} R package, available on CRAN.

The choice of $\alpha_k$ in \eqref{eq:hdrda-cov} is one of convenience and
allows the flexibility of various covariance-matrix estimators proposed in the
literature. In practice, we generally are not interested in estimating
$\alpha_k$ as it would incorporate $K$ additional tuning parameters to estimate
via cross-validation, which is counterproductive to our goal of improving the
\emph{RDA} classifier's efficiency. By setting $\alpha_k$ to an appropriate
constant, the \emph{HDRDA} covariance-matrix estimator includes or resembles a
large family of estimators. Notice that if $\alpha_k = 1$ and $\lambda = 1$,
\eqref{eq:hdrda-cov} is equivalent to the standard ridge-like covariance-matrix
estimator in \eqref{eq:ridge-estimator}, and various estimators proposed in the
literature can be obtained by selecting $\gamma$ accordingly. For instance, with
$\gamma = \tr\{ \widehat{\bm \Sigma} \} / \min(N, p)$, we obtain the estimator
from \cite{Srivastava:2007ww}.

Perhaps the most well-known of the ridge-like estimators is from
\cite{Friedman:1989tm}, who proposed regularized discriminant analysis (RDA) by
incorporating a weighted average of the sample covariance matrix $\widehat{ \bm
  \Sigma}_k$ and the pooled sample covariance matrix $\widehat{\bm \Sigma}$ to
estimate $\bm \Sigma_k$ with
\begin{align}
    \widehat{\bm \Sigma}_k(\lambda) = n_k^{-1}(\lambda) \bm S_k(\lambda) \quad
    (k =
      1, \ldots, K),\label{eq:sig-lambda}
      \end{align}
where $\lambda \in [0, 1]$, $n_k(\lambda) = (1 - \lambda) n_k + \lambda N$, $
\bm S_k = n_k \widehat{\bm \Sigma}_k$, $\bm S = \sum_{k=1}^K \bm S_k$, and $ \bm
S_k(\lambda) = (1 - \lambda) \bm S_k + \lambda \bm S$. If $\lambda = 0$,
\eqref{eq:sig-lambda} is equal to $\widehat{\bm \Sigma}_k$ in \eqref{eq:qda}.
Alternatively, if $\lambda = 1$, then \eqref{eq:sig-lambda} reduces to
\eqref{eq:pooled-cov}. To further improve the estimation of $\bm \Sigma_k$ and
to stabilize the inverse of \eqref{eq:sig-lambda}, \cite{Friedman:1989tm} has
proposed the biased covariance-matrix estimator
\begin{align}
          \widehat{\bm \Sigma}_k(\lambda, \gamma) = (1 - \gamma) \widehat{\bm
            \Sigma}_k(\lambda) + \gamma \frac{\tr\{\widehat{\bm
              \Sigma}_k(\lambda)\}}{p} \bm I_p,\label{eq:sig-rda}
          \end{align}
where $\gamma \in [0, 1]$ is a regularization parameter that controls the
shrinkage of \eqref{eq:sig-rda} towards $\bm I_p$ and is weighted by the average
of the eigenvalues of \eqref{eq:sig-lambda}. Finally, the RDA classifier is
established by substituting \eqref{eq:sig-rda} into \eqref{eq:qda}.

Our proposed \emph{HDRDA} classifier resembles \eqref{eq:sig-rda} but improves
the interpretation of the pooling parameter $\lambda$ as discussed
above. Moreover, as we discuss below, our formulation of \emph{HDRDA} classifier
lends itself to rapid estimation in high-dimensional classification.

@@

As we can see, we are able to avoid the time-consuming inverses and determinants
of $p \times p$ covariance matrices in \eqref{eq:hdrda} and instead calculating
these same operations on $\bm W_k \in \mathbb{R}_{q \times q}$ in
\eqref{eq:hdrda-decomposed}. The substantial computational improvements arise
because our proposed classifier in \eqref{eq:hdrda} is invariant to the term
$\bm U_2$, thus yielding an equivalent classifier in \eqref{eq:hdrda-decomposed}
with a substantial reduction in computational complexity. Here, we demonstrate
that the computational efficiency in calculating the inverse and determinant of
$\bm W_k$ can be further improved via standard matrix operations by showing that
the inverses and determinants of $\bm W_k$ can be performed on matrices of size
$n_k \times n_k$.

>>>>

\cite{Friedman:1989tm} proposed the popular regularized discriminant analysis
(RDA) classifier that utilizes a biased covariance-matrix estimator that
partially pools the sample covariance matrices from linear and quadratic
disriminant analysis and shrinks the resulting estimator towards a scaled
identity matrix. The RDA classifier's two tuning parameters are typically
estimated via a computationally burdensome cross-validation procedure using a
grid search. We formulate a new RDA classifier for the small-sample,
high-dimensional setting and then show that the classification decision rule is
equivalent to a classifier in a subspace having a much lower dimension. As a
result, the utilization of the dimension-reduction step yields a substantial
reduction in computation during model selection. Also, our parameterization
offers interpretability previously lacking with the RDA classifier. We
demonstrate that our proposed classifier is often superior to several recently
proposed sparse and regularized classifiers in terms of classification accuracy
with three artificial and six real high-dimensional data sets. Finally, we
provide an implementation of our proposed classifier in the {\tt sparsediscrim}
R package, which is available on CRAN.

@@

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers}
employed in a variety of disciplines, such as medical research, facial
recognition, remote sensing, and chemical analysis
\citep*{Ching:2012fu,Li:2012ev,Zhang:2010va}. Despite the advances in
classification to facilitate technological advancements, well-established
classifiers often degrade in accuracy when applied to modern, high-dimensional
data sets if the number of features $p$ exceeds the total sample size $N$, due
to the \emph{curse of dimensionality} \citep{Bellman:1961tn}. For instance,
linear discriminant analysis (\emph{LDA}) and quadratic discriminant analysis
(\emph{QDA}) are popular because of their simplicity, speed of computation, and
interpretability. However, both methods are incalculable when $p > N$ because
the class sample covariance matrices are singular \citep*{Bouveyron:2007gx}.

@@

Here, we present the high-dimensional \emph{RDA} (\emph{HDRDA}) classifier,
which is intended for the $p > N$ case and builds on the classic \emph{RDA}
classifier proposed by \cite{Friedman:1989tm}. In this paper, we study the
classification performance of the \emph{HDRDA} classifier on six real
high-dimensional data sets along with a simulation design that generalizes the
experiments initially conducted by \cite{Guo:2007te}. We demonstrate that the
\emph{HDRDA} classifier often attains superior classification accuracy to
several recent classifiers designed for small-sample, high-dimensional data from
\cite*{Tong:2012hw}, \cite{Witten:2011kc}, \cite*{Pang:2009ik}, and
\cite{Guo:2007te}. We also include as a benchmark the random forest from
\cite{Breiman:2001fb} because \cite*{FernandezDelgado:2014ul} have concluded
that the random forest is often superior to other classifiers in benchmark
studies. We show that our proposed classifier is competitive and often
outperforms the random forest in terms of classification accuracy in the
small-sample, high-dimensional setting.

Our proposed \emph{HDRDA} classifier employs a biased covariance-matrix
estimator that partially pools the individual sample covariance matrices from
the \emph{QDA} classifier with the pooled sample covariance matrix from the
\emph{LDA} classifier. We then shrink the resulting covariance-matrix estimator
towards a scaled identity matrix to ensure positive definiteness. By
reparameterizing the \emph{RDA} classifier similar to \cite*{Hastie:2008dt} and
\cite{Halbe:2007ea}, we show that the pooling parameter in the \emph{HDRDA}
classifier determines the contribution of each training observation to the
estimation of each class covariance matrix, enabling interpretability that has
been previously lacking with the \emph{RDA} classifier \citep{Bensmail:1996fw}.

Next, we achieve a substantial reduction in dimension that greatly reduces the
computational effort during model selection when $p \gg N$ by applying reasoning
similar to \cite{Ye:2006tq}. We establish that the matrix operations
corresponding to the null space of the pooled sample covariance matrix are
redundant and can be discarded from the \emph{HDRDA} decision rule without loss
of classificatory information. Moreover, because the \emph{HDRDA} decision rule
is invariant to adjustments to the zero eigenvalues, the decision rule in the
original feature space is equivalent to a decision rule in a lower dimension,
such that matrix inverses and determinants of relatively small matrices can be
rapidly computed. Also, we provide an efficient algorithm along with pseudocode
to estimate the \emph{HDRDA} classifier's tuning parameters in a grid search via
cross-validation. Finally, we show that several shrinkage methods that are
special cases of the \emph{HDRDA} classifier have no effect on the zero
eigenvalues of the covariance-matrix estimators. Such methods include work from
\cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

The remainder of this paper is organized as follows. In Section
\ref{sec:preliminaries} we present the discrimination problem and necessary
notation to describe our contributions. In Section \ref{sec:rda} we present the
\emph{HDRDA} classifier along with its interpretation, properties, and
computationally efficient model-selection procedure. In Section \ref{sec:sims},
we describe our simulation studies of artificial and real data sets and also
describes the experimental results. We conclude with a brief discussion in
Section \ref{sec:discussion}.

@@

The smallest eigenvalues of $\widehat{\bm \Sigma}_k$ and the directions
associated with their eigenvectors can highly influence the classifier in
\eqref{eq:qda}. In fact, the eigenvalues of $\widehat{ \bm \Sigma}_k$ are
well-known to be biased if $p \ge n_k$ such that the smallest eigenvalues are
underestimated \citep{Seber:2004uh}. Moreover, if $p > n_k$, then
rank$(\widehat{\bm \Sigma}_k) \le n_k$, which implies that at least $p - n_k$
eigenvalues of $\widehat{\bm \Sigma}_k$ are zero. Furthermore, although more
feature information is available to discriminate among the $K$ classes, if $p >
n_k$, \eqref{eq:qda} is incalculable because $\widehat{\bm \Sigma}_k^{-1}$ does
not exist.

Several regularization methods, such as the methods considered by
\cite{Xu:2009fl}, \cite{Guo:2007te}, and \cite{Mkhadri:1995jp} have been
proposed in the literature to adjust the eigenvalues of $\widehat{\bm \Sigma}_k$
so that \eqref{eq:qda} is calculable and provide reduced variability for
$\widehat{\bm \Sigma}_k^{-1}$. A common form of the covariance-matrix regularization applies
  a shrinkage factor $\gamma > 0$, so that
\begin{align}
	\widehat{\bm \Sigma}_k(\gamma) = \widehat{\bm \Sigma}_k + \gamma \bm I_p, \label{eq:ridge-estimator}
\end{align}
similar to a method employed in ridge regression \citep{Hoerl:1970cd}. Equation
\eqref{eq:ridge-estimator} effectively \emph{shrinks} the sample covariance
matrix $\widehat{\bm \Sigma}_k$ toward $\bm I_p$, thereby increasing the
eigenvalues of $\widehat{\bm \Sigma}_k$ by $\gamma$. Specifically, the zero
eigenvalues are replaced with $\gamma$, so that \eqref{eq:ridge-estimator} is
positive definite. For additional covariance-matrix regularization methods, see
\cite{Ramey:2013ji}, \cite{Xu:2009fl}, and \cite{Ye:2009gd}.

@@

Here, we define the \emph{HDRDA} classifier by first formulating the
covariance-matrix estimator $\widehat{\bm \Sigma}_k(\lambda)$ and demonstrating
its clear interpretation as a linear combination of the crossproducts of the
training observations centered by their respective class sample means. We define
the convex combination
\begin{align}
  \widehat{\bm \Sigma}_k(\lambda) \coloneqq (1 - \lambda) \widehat{\bm \Sigma}_k + \lambda
  \widehat{\bm \Sigma}, \quad k = 1, \ldots, K,\label{eq:sig-lambda-alternative}
\end{align}
where $\lambda \in [0, 1]$ is the \emph{pooling} parameter. By rewriting
\eqref{eq:sig-lambda-alternative} in terms of the observations $\bm x_i$, $i =
1, \ldots, N$, each centered by its class sample mean, we attain a clear
interpretation of $\widehat{\bm \Sigma}_k(\lambda)$. That is,
\begin{align}
	\widehat{\bm \Sigma}_k(\lambda)
  &= \left( 1 - \lambda + \frac{\lambda n_k}{N} \right) \widehat{\bm \Sigma}_k +  \frac{\lambda}{N} \sum_{\substack{k' = 1\\k' \ne k}}^K n_{k'} \widehat{\bm \Sigma}_{k'} \nonumber \\
	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right)\sum_{i=1}^N
  I(y_i = k) \bm x_i \bm x_i^{T} +  \frac{\lambda}{N} \sum_{i=1}^N I(y_i \ne k)
  \bm x_i \bm x_i^{T} \nonumber \\
	&= \sum_{i=1}^N c_{ik}(\lambda) \bm x_i \bm x_i^{T},\label{eq:sig-lambda-alternative2}
\end{align}
where $c_{ik}(\lambda) = \lambda N^{-1} + (1 - \lambda)n_k^{-1}I(y_i = k)$.
From \eqref{eq:sig-lambda-alternative2}, we see that $\lambda$ weights the
contribution of each of the $N$ observations in estimating $\bm \Sigma_k$ from
all $K$ classes rather than using only the $n_k$ observations from a single
class. As a result, we can interpret \eqref{eq:sig-lambda-alternative2} as a
covariance-matrix estimator that borrows from $\widehat{\bm \Sigma}$ in
\eqref{eq:pooled-cov} to estimate $ \bm \Sigma_k$.

In Figure \ref{fig:hdrda-contours} we plot the contours of five multivariate
normal populations for $\lambda = 0$ with unequal covariance matrices. As
$\lambda$ approaches 1, the contours become more similar, resulting in identical
contours for $\lambda = 1$. Below, we show that the pooling operation is
advantageous in increasing the rank of each $\widehat{\bm \Sigma}_k(\lambda)$
from rank$(\widehat{\bm \Sigma}_k)$ to rank$(\widehat{\bm \Sigma})$ for $0 <
\lambda \le 1$. Notice that if $\lambda = 0$, then the observations from the
remaining $K - 1$ classes do not contribute to the estimation of $\bm \Sigma_k$,
corresponding to $\widehat{\bm \Sigma}_k$. Furthermore, if $\lambda = 1$, the
weights $c_{ik}(\lambda)$ in \eqref{eq:sig-lambda-alternative2} reduce to $1/N$,
corresponding to $\widehat{\bm \Sigma}$. For brevity, when $\lambda = 1$, we
define $\bm X = [\sqrt{c_{1k}(1)} \bm x_1^{T}, \ldots, \sqrt{c_{Nk}(1)} \bm
  x_N^{T}]^{T}$ such that $\widehat{\bm \Sigma} = N^{-1} \bm X^{T} \bm
X$. Similarly, for $\lambda = 0$, we define $\bm X_k = [\sqrt{c_{1k}(0)} \bm
  x_1^{T}, \ldots, \sqrt{c_{Nk}(0)} \bm x_N^{T}]^{T}$ such that $\widehat{\bm
  \Sigma}_k = n_k^{-1} \bm X_k^{T} \bm X_k$.

@@

As we have discussed above, several eigenvalue adjustment methods have been
proposed to increase eigenvalues (approximately) equal to 0. To further improve
the estimation of $\bm \Sigma_k$ and to stabilize the estimator's inverse, we
define the eigenvalue adjustment of \eqref{eq:sig-lambda-alternative} as
\begin{align}
	\tilde{\bm \Sigma}_k \coloneqq \alpha_k \widehat{\bm \Sigma}_k(\lambda) + \gamma \bm I_p,\label{eq:hdrda-cov}
\end{align}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. Thus, the \emph{pooling} parameter $\lambda$ controls the amount of
estimation information borrowed from $\widehat{\bm \Sigma}$ to estimate $\bm
\Sigma_k$, and the \emph{shrinkage} parameter $\gamma$ determines the degree of
eigenvalue shrinkage. The choice of $\alpha_k$ allows for a flexible formulation
of covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1,
\ldots, K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}.
Similarly, if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form
comparable to the \emph{RDA} classifier from \cite{Friedman:1989tm}.
Substituting \eqref{eq:hdrda-cov} into \eqref{eq:qda}, we define the
\emph{HDRDA} classifier
\begin{align}
	D_{HDRDA}(\bm x) = \argmin_{k}  (\bm x - \xbar_k)^{T}\tilde{\bm \Sigma}_k^{+}(\bm x - \xbar_k)  + \log |\tilde{\bm \Sigma}_k|. \label{eq:hdrda}
\end{align}
For $\gamma > 0$, $\tilde{\bm \Sigma}_k$ is nonsingular such that $\tilde{\bm
  \Sigma}_k^{-1}$ can be substituted for $\tilde{\bm \Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\bm \Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\bm
  \Sigma}_k$. Following \cite{Friedman:1989tm}, we select $\lambda$ and $\gamma$
from a grid of candidate models via cross-validation. We provide an
implementation of \eqref{eq:hdrda} in the {\tt hdrda} function contained in the
{\tt sparsediscrim} R package, which is available on CRAN.

The choice of $\alpha_k$ in \eqref{eq:hdrda-cov} is one of convenience and
allows the flexibility of various covariance-matrix estimators proposed in the
literature. In practice, we generally are not interested in estimating
$\alpha_k$ as it would incorporate $K$ additional tuning parameters to estimate
via cross-validation, which is counterproductive to our goal of improving the
\emph{RDA} classifier's computational efficiency. By setting $\alpha_k$ to an
appropriate constant, the \emph{HDRDA} covariance-matrix estimator includes or
resembles a large family of estimators. Notice that if $\alpha_k = 1$ and
$\lambda = 1$, \eqref{eq:hdrda-cov} is equivalent to the standard ridge-like
covariance-matrix estimator in \eqref{eq:ridge-estimator}, and various
estimators proposed in the literature can be obtained by selecting $\gamma$
accordingly. For instance, with $\gamma = \tr\{ \widehat{\bm \Sigma} \} /
\min(N, p)$, we obtain the estimator from \cite{Srivastava:2007ww}.

Perhaps the most well-known of the ridge-like estimators is from
\cite{Friedman:1989tm}, who proposed regularized discriminant analysis (RDA) by
incorporating a weighted average of the sample covariance matrix $\widehat{ \bm
  \Sigma}_k$ and the pooled sample covariance matrix $\widehat{\bm \Sigma}$ to
estimate $\bm \Sigma_k$ with
\begin{align}
    \widehat{\bm \Sigma}_k(\lambda) = n_k^{-1}(\lambda) \bm S_k(\lambda) \quad
    (k =
      1, \ldots, K),\label{eq:sig-lambda}
      \end{align}
where $\lambda \in [0, 1]$, $n_k(\lambda) = (1 - \lambda) n_k + \lambda N$, $
\bm S_k = n_k \widehat{\bm \Sigma}_k$, $\bm S = \sum_{k=1}^K \bm S_k$, and $ \bm
S_k(\lambda) = (1 - \lambda) \bm S_k + \lambda \bm S$. If $\lambda = 0$,
\eqref{eq:sig-lambda} is equal to $\widehat{\bm \Sigma}_k$ in \eqref{eq:qda}.
Alternatively, if $\lambda = 1$, then \eqref{eq:sig-lambda} reduces to
\eqref{eq:pooled-cov}. To further improve the estimation of $\bm \Sigma_k$ and
to stabilize the inverse of \eqref{eq:sig-lambda}, \cite{Friedman:1989tm}
proposed the biased covariance-matrix estimator
\begin{align}
          \widehat{\bm \Sigma}_k(\lambda, \gamma) = (1 - \gamma) \widehat{\bm
            \Sigma}_k(\lambda) + \gamma \frac{\tr\{\widehat{\bm
              \Sigma}_k(\lambda)\}}{p} \bm I_p,\label{eq:sig-rda}
          \end{align}
where $\gamma \in [0, 1]$ is a regularization parameter that controls the
shrinkage of \eqref{eq:sig-rda} towards $\bm I_p$ and is weighted by the average
of the eigenvalues of \eqref{eq:sig-lambda}. Finally, the RDA classifier is
established by substituting \eqref{eq:sig-rda} into \eqref{eq:qda}.

Our proposed \emph{HDRDA} classifier resembles \eqref{eq:sig-rda} but yields an
interpretation of the pooling parameter $\lambda$ as discussed above. Moreover,
as we discuss below, our formulation of \emph{HDRDA} classifier lends itself to
rapid estimation in high-dimensional classification.

@@

As we can see, we are able to avoid the time-consuming inverses and determinants
of $p \times p$ covariance matrices in \eqref{eq:hdrda} and instead calculate
these same operations on $\bm W_k \in \mathbb{R}_{q \times q}$ in
\eqref{eq:hdrda-decomposed}. The substantial computational improvements arise
because our proposed classifier in \eqref{eq:hdrda} is invariant to the term
$\bm U_2$, thus yielding an equivalent classifier in \eqref{eq:hdrda-decomposed}
with a substantial reduction in computational complexity. Here, we demonstrate
that the computational efficiency in calculating the inverse and determinant of
$\bm W_k$ can be further improved via standard matrix operations by showing that
the inverses and determinants of $\bm W_k$ can be performed on matrices of size
$n_k \times n_k$.
